---
title: "Mend PA"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true

---

# Data Cleaning
Take the participant's "subjectdaystats" file and delete rows not part of the participant's wear days. Rename the file to XXXdaystats. Upload this CSV file to R. Assign the cleaned data file to a new object called "activity_clean". Create a new column called "Day" that is 1-15. Change the date format to "Month","Day", "Year". 
Make new columns for each activity level ie. look for "sedentary, count" and find the digits directly after that. Filter so only it only shows Days 2-15 

```{r echo=FALSE, message = FALSE, results='hide'}
### hide r code and packages




library(tidyverse)
library(dplyr)
library(mice)
library(knitr)

library(stringr)
setwd(dir = "/Users/jade/Desktop/TC/MEND")
data <- read.csv(file = "0009daystats.csv")

# clean up the data so there is a  column for each activity level 
# example: look for "sedentary, count" and find the digits directly after 
#create a new column called "Day" that is 1-15 
#change the date format so it is Month,Day,Year 

activity_clean <- data %>%
group_by(Subject) %>%
  mutate(
    Date = format(as.Date(Date), "%m-%d-%Y"),
    Day = 1:n(),
    Sedentary = as.numeric(str_extract(WearFilteredCutPoints, '(?<="Sedentary","Count":)\\d+')),
    Light     = as.numeric(str_extract(WearFilteredCutPoints, '(?<="Light","Count":)\\d+')),
    Moderate  = as.numeric(str_extract(WearFilteredCutPoints, '(?<="Moderate","Count":)\\d+')),
    Vigorous  = as.numeric(str_extract(WearFilteredCutPoints, '(?<="Vigorous","Count":)\\d+')),
    `Very Vigorous`  = as.numeric(str_extract(WearFilteredCutPoints, '(?<="Very Vigorous","Count":)\\d+'))
  
  ) %>%
  filter(Day >= 2 & Day <= 15)
#just include days 2-15 
```
# Data Transformation and Summary 
Reshape data to long format and calculate average counts and percentages for each activity level. For each activity level, take the average counts and turn it into a percentage so it shows the percentage of time the participant spent in each activity level 
```{r echo=FALSE, message = FALSE, results='hide'}
activity_chart <- activity_clean %>%
  group_by(Subject)%>%
  select(Day, Sedentary, Light, Moderate, Vigorous,`Very Vigorous`)  %>%
  pivot_longer(cols = c(Sedentary,Light, Moderate,Vigorous,`Very Vigorous`),
               names_to = "Activity",
               values_to = "Count")
activity_chart


activity_summary <- activity_chart %>%
  group_by(Subject,Activity) %>%
  summarise(Average = mean(Count, na.rm = TRUE))%>%
  mutate(Percent = round(Average / sum(Average) * 100, 1))%>%
  mutate(Activity = factor(Activity, levels = c("Sedentary","Light", "Moderate", "Vigorous","Very Vigorous")))







```
# Average Activity Distribution (Pie Chart)
Create a pie chart showing the activity intensity distribution 

```{r,fig.show="hold", fig.width=10,fig.height=5, echo=FALSE, message=FALSE}

ggplot(activity_summary, aes(x = "", y = Average, fill = Activity)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  geom_text(
    aes(label = ifelse(Percent >0, paste0(Percent,"%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black",
    size = 5
  ) +
  scale_fill_manual(values = c(
    "Sedentary" = "#00FF00",
    "Light"    = "#FFD700",
    "Moderate" = "#FF6EB4",
    "Vigorous" = "#8da0cb",
    "Very Vigorous" = "#9370DB"
    
  ))+
  labs(
    title = "Average Activity Distribution across 14 Days",
    subtitle = "000X",
    fill = "Activity Level"
  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic") 
  )

```
**Figure 1** displaying the activity intensity distribution for the participant across 14 days.

# Daily Steps Plot
Create separate plots for Week 1( Day 2-7) and Week 2( Day 8-15) and add a shaded confidence interval 
```{r,fig.show="hold", fig.width=10,fig.height=5, echo=FALSE, message=FALSE}
ggplot(filter(activity_clean, Day >= 2 & Day <= 7),
       aes(x = Day, y = WearFilteredSteps)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") + 
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, face = "italic"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
  ) +
  labs(
    title = "Daily Steps",
    subtitle = "000X",
    x = "Week 1",
    y = "Steps"
  )

ggplot(filter(activity_clean, Day >= 8 & Day <= 15),
       aes(x = Day, y = WearFilteredSteps)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") + 
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, face = "italic"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
  ) +
  labs(
    title = "Daily Steps",
    subtitle = "000X",
    x = "Week 2",
    y = "Steps"
  )
```
**Figure 2** displays the daily steps for the participant across 14 days 

