---
title: "MEND-HD Master Dataset Construction"
author: "Meghan"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This document constructs the harmonized 14-day multimodal master dataset for the MEND-HD project.

All files are assumed to be located in the same directory as this `.Rmd` file.

---

# Load Required Packages

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(openxlsx)
```

---

# Load Dates of Participants

```{r}
date_ranges <- read_excel("Date_RangesMF0303.xlsx")

date_ranges$SUBJECT <- sprintf("%04d", as.numeric(date_ranges$SUBJECT))
date_ranges$START_DATE <- as.Date(date_ranges$START_DATE)
date_ranges$END_DATE   <- as.Date(date_ranges$END_DATE)
```

---

# Build Demographic Backbone

```{r}
demo_raw <- read.csv(
  "MENDHD_DATA_LABELS_2026-03-02_1609.csv",
  stringsAsFactors = FALSE
)

# Remove Care Partner and Screen Fails (including 0005)
demo_raw <- demo_raw %>%
  filter(!grepl("c$", Record.ID))

baseline_ids <- demo_raw %>%
  filter(Event.Name == "Visit 3 (Baseline)") %>%
  distinct(Record.ID)

demo_raw <- demo_raw %>%
  filter(Record.ID %in% baseline_ids$Record.ID)

demo_raw <- demo_raw %>%
  filter(Record.ID != "0005")

# Filter for Age and Sex
demo_visit1 <- demo_raw %>%
  filter(Event.Name == "Visit 1 (Screening)") %>%
  select(Record.ID, Age, Sex)

# Filter for Stage
demo_visit2 <- demo_raw %>%
  filter(Event.Name == "Visit 2 (Assessment)") %>%
  select(Record.ID, Investigators.assessed.Stage.)

# Create Master Demographics Table
demo_master <- left_join(
  demo_visit1,
  demo_visit2,
  by = "Record.ID"
)

# Label Controls
demo_master <- demo_master %>%
  mutate(
    Investigators.assessed.Stage. =
      ifelse(
        is.na(Investigators.assessed.Stage.) |
          Investigators.assessed.Stage. == "",
        "Control",
        Investigators.assessed.Stage.
      )
  ) %>%
  rename(SUBJECT = Record.ID)
```

---

# Daily Wear Metrics

```{r}
daily_raw <- read.csv("subjectdaystats.csv", stringsAsFactors = FALSE)

daily_raw <- daily_raw %>%
  filter(!grepl("^test", Subject, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(Subject)),
    DATE = as.Date(substr(Date, 1, 10))
  )

daily_joined <- left_join(daily_raw, date_ranges, by = "SUBJECT")

daily_filtered <- daily_joined %>%
  filter(DATE >= START_DATE & DATE <= END_DATE)

daily_final <- daily_filtered %>%
  select(SUBJECT, DATE, WearMinutes, Steps, MVPA)

master_data <- left_join(demo_master, daily_final, by = "SUBJECT")
```

---

# Activity Intensity

```{r}
activity_raw <- read.csv(
  "activity_23_5508340f-66b7-4f82-b427-ad8166649969_activitydaily29.csv",
  stringsAsFactors = FALSE
)

activity_raw <- activity_raw %>%
  filter(!grepl("^test", SUBJECT, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(SUBJECT)),
    DATE = as.Date(substr(DATE, 1, 10))
  )

activity_joined <- left_join(activity_raw, date_ranges, by = "SUBJECT")

activity_filtered <- activity_joined %>%
  filter(DATE >= START_DATE & DATE <= END_DATE)

activity_final <- activity_filtered %>%
  select(
    SUBJECT,
    DATE,
    AWAKE_WEAR_FILTERED_SEDENTARY_BEHAVIOR,
    AWAKE_WEAR_FILTERED_LIGHT_ACTIVITY,
    AWAKE_WEAR_FILTERED_MODERATE_ACTIVITY,
    AWAKE_WEAR_FILTERED_VIGOROUS_ACTIVITY
  )

master_data <- left_join(master_data, activity_final,
                         by = c("SUBJECT", "DATE"))
```

---

# Gait Metrics

```{r}
gait_raw <- read.csv(
  "activity_23_5508340f-66b7-4f82-b427-ad8166649969_gaitdailymetricscsem19.csv",
  stringsAsFactors = FALSE
)

gait_raw <- gait_raw %>%
  filter(!grepl("^test", SUBJECT, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(SUBJECT)),
    DATE = as.Date(substr(DATE, 1, 10))
  )

gait_joined <- left_join(gait_raw, date_ranges, by = "SUBJECT")

gait_filtered <- gait_joined %>%
  filter(DATE >= START_DATE & DATE <= END_DATE)

gait_final <- gait_filtered %>%
  select(SUBJECT, DATE, TOTAL_DISTANCE, CADENCE_MEAN, SPEED_MEAN)

master_data <- left_join(master_data, gait_final,
                         by = c("SUBJECT", "DATE"))
```

---

# Heart Rate Metrics

```{r}
hr_raw <- read.csv(
  "heartrate_23_34b3ec0c-1105-4f14-8dc3-d1b18d5f23f3_heartratedays13.csv",
  stringsAsFactors = FALSE
)

hr_raw <- hr_raw %>%
  filter(!grepl("^test", SUBJECT, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(SUBJECT)),
    DATE = as.Date(substr(DATE, 1, 10))
  )

hr_joined <- left_join(hr_raw, date_ranges, by = "SUBJECT")

hr_filtered <- hr_joined %>%
  filter(DATE >= START_DATE & DATE <= END_DATE)

hr_final <- hr_filtered %>%
  select(SUBJECT, DATE, HR_MEAN_24H, HR_STD_24H)

master_data <- left_join(master_data, hr_final,
                         by = c("SUBJECT", "DATE"))
```

---

# Heart Rate Variability (with Transformations)

```{r}
hrv_raw <- read.csv(
  "heartrate_23_34b3ec0c-1105-4f14-8dc3-d1b18d5f23f3_heartratevariabilitydays12.csv",
  stringsAsFactors = FALSE
)

hrv_raw <- hrv_raw %>%
  filter(!grepl("^test", SUBJECT, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(SUBJECT)),
    DATE = as.Date(substr(DATE, 1, 10))
  )

hrv_joined <- left_join(hrv_raw, date_ranges, by = "SUBJECT")

hrv_filtered <- hrv_joined %>%
  filter(DATE >= START_DATE & DATE <= END_DATE) %>%
  mutate(
    RMSSD_90PCT_24H = as.numeric(RMSSD_90PCT_24H),
    SDNN_90PCT_24H  = as.numeric(SDNN_90PCT_24H),
    SDSD_90PCT_24H  = as.numeric(SDSD_90PCT_24H),
    PNN20_90PCT_24H = as.numeric(PNN20_90PCT_24H),
    LF_POWER_90PCT_24H = as.numeric(LF_POWER_90PCT_24H),
    HF_POWER_90PCT_24H = as.numeric(HF_POWER_90PCT_24H),
    LF_HF_Ratio = ifelse(HF_POWER_90PCT_24H == 0,
                         NA,
                         LF_POWER_90PCT_24H / HF_POWER_90PCT_24H)
  )

hrv_final <- hrv_filtered %>%
  select(
    SUBJECT,
    DATE,
    RMSSD_90PCT_24H,
    SDNN_90PCT_24H,
    SDSD_90PCT_24H,
    PNN20_90PCT_24H,
    LF_POWER_90PCT_24H,
    HF_POWER_90PCT_24H,
    LF_HF_Ratio
  )

master_data <- left_join(master_data, hrv_final,
                         by = c("SUBJECT", "DATE"))
```

---

# Sleep (Night-Based Transformation)

```{r}
sleep_raw <- read.csv(
  "subjectsleepperiodmetrics.csv",
  stringsAsFactors = FALSE
)

sleep_raw <- sleep_raw %>%
  filter(!grepl("^test", Subject, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(Subject)),
    InBedTime = as.POSIXct(InBedTime,
                           format = "%Y-%m-%dT%H:%M:%S",
                           tz = "UTC")
  )

sleep_raw <- sleep_raw %>%
  mutate(
    hour = as.numeric(format(InBedTime, "%H")),
    NightDate = case_when(
      hour >= 17 ~ as.Date(InBedTime),
      hour < 8   ~ as.Date(InBedTime) - 1,
      TRUE       ~ NA_Date_
    )
  ) %>%
  filter(!is.na(NightDate))

sleep_nightly <- sleep_raw %>%
  group_by(SUBJECT, NightDate) %>%
  summarise(
    WakeAfterOnsetInMinutes =
      sum(as.numeric(WakeAfterOnsetInMinutes), na.rm = TRUE),
    AwakeningCount =
      sum(as.numeric(AwakeningCount), na.rm = TRUE),
    .groups = "drop"
  )

sleep_filtered <- left_join(sleep_nightly, date_ranges,
                            by = "SUBJECT") %>%
  filter(NightDate >= START_DATE &
           NightDate <= END_DATE) %>%
  arrange(SUBJECT, NightDate) %>%
  group_by(SUBJECT) %>%
  mutate(NIGHT = row_number()) %>%
  ungroup()

sleep_final <- sleep_filtered %>%
  rename(DATE = NightDate) %>%
  select(SUBJECT, DATE, NIGHT,
         WakeAfterOnsetInMinutes, AwakeningCount)

master_data <- left_join(master_data, sleep_final,
                         by = c("SUBJECT", "DATE"))
```

---

# Sleep Stages

```{r}
sleepmetrics_raw <- read.csv(
  "sleepmetricsdacnn_23_c5669506-b1d1-4b2a-a13b-ed8b3cf518da_sleepmetrics4.csv",
  stringsAsFactors = FALSE
)

sleepmetrics_raw <- sleepmetrics_raw %>%
  filter(!grepl("^test", SUBJECT, ignore.case = TRUE)) %>%
  mutate(
    SUBJECT = sprintf("%04d", as.numeric(SUBJECT)),
    DATE = as.Date(substr(DATE, 1, 10))
  )

sleepmetrics_filtered <- left_join(sleepmetrics_raw,
                                   date_ranges,
                                   by = "SUBJECT") %>%
  filter(DATE >= START_DATE & DATE <= END_DATE)

sleepmetrics_final <- sleepmetrics_filtered %>%
  select(
    SUBJECT,
    DATE,
    TOTAL_SLEEP_TIME_MINUTES,
    LIGHT_SLEEP_TIME_MINUTES,
    DEEP_SLEEP_TIME_MINUTES,
    REM_SLEEP_TIME_MINUTES,
    NREM_SLEEP_TIME_MINUTES
  )

master_data <- left_join(master_data, sleepmetrics_final,
                         by = c("SUBJECT", "DATE"))
```

---

# Final Export

```{r}
write.xlsx(
  master_data,
  file = "Master_Data_03_03_26.xlsx",
  rowNames = FALSE
)
```
